---
title: "Multi-plates input for"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drugseq_multiplates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

As all other vignettes focus on getting data from one single plate. In this vignette, we demonstrate how to read in multiple plates at a time, in a situation like combining compounds or replicates from different plates. We're going to use DRUGseq data as an example, as each batch has three replicate plates. 


# DRUGseq data

The DRUGseq dataset is a large-scale drug screening dataset that includes a large set of small molecules (N = 4,343) tested on U2OS cells. The dataset is part of the MoABox project, which aims to understand the mechanisms of action of different drugs.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r, include=FALSE}
devtools::load_all("/home/rstudio/macpie/macpie/")
```


```{r setup}
library(macpie)
library(tibble)
library(stringr)
library(pheatmap)
library(ggiraph)
library(tidyseurat)
library(dplyr)
library(purrr)
```





## Metadata and data

Three metadata files in Novartis_drugseq_U2OS_MoABox/ on [DRUGseq Github](https://github.com/Novartis/DRUG-seq/tree/main/data/Novartis_drugseq_U2OS_MoABox) and data can be downloaded from this ZENODO link: [Novartis/DRUG-seq U2OS MoABox Dataset Creators](https://zenodo.org/records/14291446).

In this vignette, we need three replicate plates from batch 24.

DRUGseq data was downloaded from *./Exp_gzip.RData* from the above Zenodo link and saved as `Exp_batch24.Rds` locally for faster loading.

Each batch is a list of plates, and each plate contains the UMI counts matrix and the corresponding metadata.


```{r}
# load("DRUGseqData/Exp_gzip.RData")
# batch24 <- Exp$`24`
# saveRDS(batch24, file = "DRUGseqData/Exp_batch24.Rds")
batch24 <- readRDS("/home/rstudio/macpie/DRUGseqData/Exp_batch24.Rds")
```


```{r}
names(batch24)
```

This batch contains 3 replicate plates, each has the UMI counts matrix and the metadata.

Now, we convert the metadata format to macpie metadata format. 


```{r}
data <- batch24$VH02012944
counts <- data$UMI.counts

metadata <- data$Annotation
head(metadata)
```


```{r}
counts[1:3,1:5]
```




```{r}

batch24_metadata <- batch24 %>% 
  map_dfr(~ {
    .x$Annotation %>%
      mutate(
        Plate_ID        = plate_barcode,
        Well_ID         = well_id,
        Barcode         = paste0(plate_barcode, "_", well_index),
        Row             = LETTERS[row],
        Column          = as.integer(col),
        Species         = "human",
        Cell_type       = cell_line_name,
        Model_type      = "MoABox",
        Time            = as.factor(hours_post_treatment),
        Unit            = "h",
        Treatment_1     = cmpd_sample_id,
        Concentration_1 = as.numeric(concentration),
        Unit_1          = unit,
        Sample_type     = if_else(well_type == "SA" & col == 24,
                                  "Positive Control",
                                  well_type)
      )
  })


batch24_metadata <- batch24_metadata%>%select(-c(batch_id, plate_barcode,plate_index, well_id,
                                                 well_index, col, row, biosample_id, external_biosample_id,
                                                 cmpd_sample_id, well_type, cell_line_name, cell_line_ncn, concentration, unit, hours_post_treatment, Sample))
```





```{r}
# create a combined UMI matrix
batch24_counts <- batch24 %>%
  map(~ {
    .x$UMI.counts %>%
      as.data.frame() %>% 
      rownames_to_column("gene_id") %>%
      separate(col = gene_id, into = c("gene_name", "chrom"), sep = ",") %>%
      mutate(gene_name = make.unique(gene_name)) %>%
      select(-chrom) %>%
      tibble::column_to_rownames(var = "gene_name") %>%
      as.matrix()
  })

binded_counts <- do.call(cbind, batch24_counts)


  
```



```{r}
as_mac <- CreateSeuratObject(counts = binded_counts, 
  min.cells = 1, 
  min.features = 1)

as_mac<- as_mac%>% inner_join(batch24_metadata, by = c(".cell"="Barcode"))
```

## Filtering


```{r}
as_mac$combined_id <- paste0(as_mac$Treatment_1,"_", as_mac$Concentration_1)

min_sample_num <- min(table(as_mac$combined_id))
mac_filtered <- filter_genes_by_expression(as_mac, group_by = "combined_id", min_counts = 10,
    min_samples = min_sample_num)

```


```{r}
p <- plot_plate_layout(as_mac, "nCount_RNA", "combined_id") + facet_wrap(~orig.ident, ncol = 1) + 
  theme(strip.text = element_text(size=10),
        axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=8),
        legend.title = element_text(size=10),
        legend.text = element_text(size=8),
        trip.background = element_blank())
girafe(ggobj = p, 
  fonts = list(sans = "sans"),
  options = list(
    opts_hover(css = "stroke:black; stroke-width:1px;")
  ))
```


