---
title: "macpie: Working with Bioconductor classes"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{macpie_bioc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```


## Overview

This vignette demonstrates how to use **macpie** with **Bioconductor-native containers**:
- Convert a **Seurat** object (the package’s demo data `mini_mac`) to **SingleCellExperiment** (SCE) and **SummarizedExperiment** (SE)
- Perform **Bioconductor-native normalization** (`scuttle::logNormCounts`)
- Call `macpie` functions either:
  - directly on SCE/SE (if you provide methods), or
  - via a **tidy metadata data frame** derived from `colData()` for functions with tidy interfaces (e.g., plate layout/QC utilities)


```{r set_wd, include=FALSE}
dir <- "/Users/liuxin/macpie_Dev/"
devtools::load_all(paste0(dir, "macpie/"))
```

```{r}
suppressPackageStartupMessages({
  library(macpie)
  library(Seurat)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(Matrix)
  library(scuttle)   # Bioconductor-native normalization
  library(dplyr)
  library(tibble)
})
```


Load demo dataset (`mini_mac`)

```{r}
data("mini_mac")  # Seurat object
class(mini_mac)
```

## Working with SingleCellExperiment

```{r}
# Convert counts/metadata/embeddings when available
sce <- as.SingleCellExperiment(mini_mac)

# Bioconductor-native normalization (adds 'logcounts' assay)
sce <- scuttle::logNormCounts(sce)

sce

```

```{r}


```

