---
title: "Check the metadata file"
output:
  rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{metadata_check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```



## Metadata import and validation

As part of the macpie package, we provide a set of functions to import and validate metadata. The metadata file should contain all the information about the samples, including sample names, experimental conditions, and other relevant variables. The metadata file has to be in a tabular format.



**Key points**: 

  - use **check_metadata** to check the metadata file for common errors
  
  - use **plot_metadata_heatmap** to visually inspect metadata integrity
  
  
### Metadata input


```{r setup}

library(macpie)

# Define project variables
project_name <- "PMMSq033"
project_metadata <- system.file("extdata/PMMSq033_metadata.csv", package = "macpie")

# Load metadata
metadata <- read_metadata(project_metadata)
metadata$Time <- as.factor(metadata$Time)
metadata$Concentration_1 <- as.factor(metadata$Concentration_1)




```


### Check column names

Check the column names of the metadata file. The metadata file should contain the following columns:

```{r}
colnames(metadata)
```

Make sure that the column names are correct and match the expected format.




### Validate metadata 

The `check_metadata` function will check the metadata file for common errors, such as missing values, incorrect data types, and other potential issues. It will also provide a summary of the metadata file, including the number of samples, the number of variables, and the number of missing values.


```{r}
# Validate metadata
validate_metadata(metadata)
```


### Visualize metadata

Lastly, let's visually inspect the large number of experimental variables, in order to correct artefacts and other metadata errors.

```{r metadata_plot, fig.width = 8, fig.height = 6}
plot_metadata_heatmap(metadata)


```


